High-Dimensional Regularized Discriminant Analysis

### 2016-12-02 reports/paper-hdrda.Rnw
Updates to the intro

In this paper, we consider the classification of small-sample, high-dimensional
data, where the number of features $p$ exceeds the training sample size $N$. In
this setting, well-established classifiers such as linear discriminant analysis
(\emph{LDA}) and quadratic discriminant analysis (\emph{QDA}) become
incalculable because the class and pooled covariance matrix estimators are
singular \citep*{Murphy:2012uq, Bouveyron:2007gx, Mkhadri:1997gy}. To improve
the estimation of the class covariance matrices estimated in the \emph{QDA}
classifier and to ensure the covariance matrix estimators are nonsingular,
\cite{Friedman:1989tm} proposed the regularized discriminant analysis
(\emph{RDA}) classifier by incorporating a weighted average of the pooled sample
covariance matrix and the class sample covariance matrix. To further improve the
estimation of the class covariance matrix and to stabilize its inverse,
\cite{Friedman:1989tm} also included a regularization component by shrinking the
covariance matrix estimator towards the identity matrix, which yields a
nonsingular estimator following the well-known ridge-regression approach of
\cite{Hoerl:1970cd}. Despite its popularity, the ``borrowing'' operation
employed in the \emph{RDA} classifier lacks interpretability
\citep{Bensmail:1996fw}, and the \emph{RDA} classifier is computationally
inefficient in that its matrix operations are cubic in the number of
features.\footnote{TODO: Add a couple of references saying RDA is slooooow.} As
a result, the \emph{RDA} classifier is impraticial for modern high-dimensional
data sets because its matrix operations are cubic in the number of features, and
the estimation of the \emph{RDA} classifier's two tuning parameters is
computationally burdensome.

@@

We provide an efficient algorithm along with pseudocode to estimate the
\emph{HDRDA} classifier's tuning parameters in a grid search via
cross-validation. Timing comparisons between our \emph{HDRDA} implementation in
the {\tt sparsediscrim} R package available on CRAN and the standard \emph{RDA}
formulation in the {\tt klaR} R package demonstrate that as the number of
features increases, the computational runtime of the \emph{HDRDA} classifier is
drastically smaller than \emph{RDA}. In fact, when $p = 1000$ and $N = 100$, the
\emph{HDRDA} classifier is \textbf{XXX} times faster on average than the
\emph{RDA} classifier.

@@

In Figure \ref{fig:timing-results}, we plot the runtime of the model selections
for both the \emph{HDRDA} and \emph{RDA} classifiers as a function of the number
of features $p$. In the left panel of Figure \ref{fig:timing-results}, we
observe that the \emph{HDRDA} classifier is substantially faster than the
\emph{RDA} classifier as $p$ increases. Figure \ref{fig:timing-results} suggests
that the usage of a matrix inverse and determinant in the {\tt klaR} R package's
discriminant function yielded training and model-selection timings that exceeded
linear growth in $p$. Because our implementation in the \emph{HDRDA} classifier
removes the inverse and determinants from the classifier, the \emph{HDRDA}
classifier is computationally more efficient than the \emph{RDA} classifier,
especially as $p$ increases. The variance in the \emph{RDA} classifier's runtime
also appeared to increase as $p$ increased. To quantify how much beyond linear
growth in $p$, we fit a regression line quadratic in $p$ to the \emph{RDA}
runtimes and plotted the fit in Figure \ref{fig:timing-results}. We attempted to
fit a cubic polynomial in $p$ to the \emph{RDA} timings, but the cubic term was
insignificant with significant quadratic and linear terms. We also fit two
polynomials to the \emph{RDA} timings with orders 2.807 and 2.376, which
correspond to the asymptotic computational runtime of two well-known algorithms
for computing a matrix inverse: the Strassen and Coppersmith-Winograd
algorithms, respectively \footnote{TODO: Citations needed}. In both cases, the
polynomials were statistically significiant at the 0.05 level. We displayed the
quadratic fit in Figure \ref{fig:timing-results} because it fit the timings best
for $p = 500$ to $p = 2000$. The fits for $2000 \le p \le 5000$ were similar.

>>>>

In this paper, we consider the classification of small-sample, high-dimensional
data, where the number of features $p$ exceeds the training sample size $N$. In
this setting, well-established classifiers such as linear discriminant analysis
(\emph{LDA}) and quadratic discriminant analysis (\emph{QDA}) become
incalculable because the class and pooled covariance matrix estimators are
singular \citep*{Murphy:2012uq, Bouveyron:2007gx, Mkhadri:1997gy}. To improve
the estimation of the class covariance matrices estimated in the \emph{QDA}
classifier and to ensure the covariance matrix estimators are nonsingular,
\cite{Friedman:1989tm} proposed the regularized discriminant analysis
(\emph{RDA}) classifier by incorporating a weighted average of the pooled sample
covariance matrix and the class sample covariance matrix. To further improve the
estimation of the class covariance matrix and to stabilize its inverse,
\cite{Friedman:1989tm} also included a regularization component by shrinking the
covariance matrix estimator towards the identity matrix, which yields a
nonsingular estimator following the well-known ridge-regression approach of
\cite{Hoerl:1970cd}. Despite its popularity, the ``borrowing'' operation
employed in the \emph{RDA} classifier lacks interpretability
\citep{Bensmail:1996fw}. Furthermore, the \emph{RDA} classifier is impractical
for high-dimensional data sets because it computes the inverse and determinant
of the covariance matrices for each class. Both matrix calculations are
computationally expensive because the number of operations grows at a polynomial
rate in the number of features. Moreover, the \emph{RDA} classifier model
selection of its two tuning parameters is computationally burdensome because the
matrix inverse and determinant of the covariance matrices across multiple
cross-validation folds for each class and for each candidate tuning-parameter
model.

@@

We provide an efficient algorithm along with pseudocode to estimate the
\emph{HDRDA} classifier's tuning parameters in a grid search via
cross-validation. Timing comparisons between our \emph{HDRDA} implementation in
the {\tt sparsediscrim} R package available on CRAN and the standard \emph{RDA}
formulation in the {\tt klaR} R package demonstrate that as the number of
features increases, the computational runtime of the \emph{HDRDA} classifier is
drastically smaller than \emph{RDA}. In fact, when $p = 5000$, we show that the
\emph{HDRDA} classifier is \Sexpr{bootstrap_summary_p5000$avg_ratio} times
faster on average than the \emph{RDA} classifier. In this scenario, the
\emph{HDRDA} classifier's training and model selection requires 3 seconds on
average, while that of the \emph{RDA} classifier requires 25 minutes on average.

@@

In Figure \ref{fig:timing-results}, we plot the runtime of the model selections
for both the \emph{HDRDA} and \emph{RDA} classifiers as a function of the number
of features $p$. In the left panel of Figure \ref{fig:timing-results}, we
observe that the \emph{HDRDA} classifier is substantially faster than the
\emph{RDA} classifier as $p$ increases. Figure \ref{fig:timing-results} suggests
that the usage of a matrix inverse and determinant in the {\tt klaR} R package's
discriminant function yielded training and model-selection timings that exceeded
linear growth in $p$. Because our implementation in the \emph{HDRDA} classifier
removes the inverse and determinants from the classifier, the \emph{HDRDA}
classifier is computationally more efficient than the \emph{RDA} classifier,
especially as $p$ increases. We contrast our implementation to the the poor
computational performance of the in the {\tt klaR} R package, which took
approximately 25 minutes on average to train the \emph{RDA} classifier on a data
set having 5,000 features. To quantify how much beyond linear growth in $p$, we
fit a regression line quadratic in $p$ to the \emph{RDA} runtimes and plotted
the fit in Figure \ref{fig:timing-results}. We attempted to fit a cubic
polynomial in $p$ to the \emph{RDA} timings, but the cubic term was
insignificant with significant quadratic and linear terms. We also fit two
polynomials to the \emph{RDA} timings with orders 2.807 and 2.376, which
correspond to the asymptotic computational runtime of two well-known algorithms
for computing a matrix inverse: the Strassen and Coppersmith-Winograd
algorithms, respectively \footnote{TODO: Citations needed}. In both cases, the
polynomials were statistically significiant at the 0.05 level. We displayed the
quadratic fit in Figure \ref{fig:timing-results} because it fit the timings best
for $p = 500$ to $p = 2000$. The fits for $2000 \le p \le 5000$ were similar.
